Weighted Model Counting with Conditional Weights for Bayesian Networks

### 2021-02-07 doc/paper2/paper.tex
New paper v1

\documentclass{article}
\pdfpagewidth=8.5in
\pdfpageheight=11in
\usepackage{ijcai21}
\usepackage{times}
\usepackage{soul}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage[utf8]{inputenc}
\usepackage[small]{caption}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{booktabs}
\urlstyle{same}

\usepackage{complexity}
\usepackage[capitalise]{cleveref}
\usepackage{amsfonts}
\usepackage{thmtools}
\usepackage{thm-restate}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage[binary-units]{siunitx}
\usepackage[inline]{enumitem}
\usepackage[table]{xcolor}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{tikz}

% the following package is optional:
%\usepackage{latexsym}

% See https://www.overleaf.com/learn/latex/theorems_and_proofs
% for a nice explanation of how to define new theorems, but keep
% in mind that the amsthm package is already included in this
% template and that you must *not* alter the styling.
\newtheorem{example}{Example}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}

\DeclareMathOperator{\im}{im}
\newcommand{\etal}{\textit{et al}.}
\Crefname{algocf}{Algorithm}{Algorithms}
\Crefname{figure}{Figure}{Figures}

@@

% Following comment is from ijcai97-submit.tex:
% The preparation of these files was supported by Schlumberger Palo Alto
% Research, AT\&T Bell Laboratories, and Morgan Kaufmann Publishers.
% Shirley Jowell, of Morgan Kaufmann Publishers, and Peter F.
% Patel-Schneider, of AT\&T Bell Laboratories collaborated on their
% preparation.

% These instructions can be modified and used in other conferences as long
% as credit to the authors and supporting agencies is retained, this notice
% is not changed, and further modification or reuse is not restricted.
% Neither Shirley Jowell nor Peter F. Patel-Schneider can be listed as
% contacts for providing assistance without their prior permission.

% To use for other conferences, change references to files and the
% conference appropriate and use other authors, contacts, publishers, and
% organizations.
% Also change the deadline and address for returning papers and the length and
% page charge instructions.
% Put where the files are available in the appropriate places.

%PDF Info Is REQUIRED.
\pdfinfo{
/TemplateVersion (IJCAI.2021.0)
}

@@

% Single author syntax
% \author{
%     Zhi-Hua Zhou
%     \affiliations
%     Nanjing University
%     \emails
%     pcchair@ijcai-21.org
% }

% Multiple author syntax (remove the single-author syntax above and the \iffalse ... \fi here)
% Check the ijcai21-multiauthor.tex file for detailed instructions
% \iffalse
% \author{
% First Author$^1$
% \and
% Second Author$^2$\and
% Third Author$^{2,3}$\And
% Fourth Author$^4$
% \affiliations
% $^1$First Affiliation\\
% $^2$Second Affiliation\\
% $^3$Third Affiliation\\
% $^4$Fourth Affiliation
% \emails
% \{first, second\}@example.com,
% third@other.example.com,
% fourth@example.com
% }
% \fi

\begin{document}

\maketitle

@@

\section{Introduction}
% TODO: mention more recent papers
% TODO: emphasise the contributions (even more)
% TODO: write ADDMC, Ace, query-dnnf, Cachet, bn2cnf, c2d in a tt font
% TODO: double-check the capitalisation of, e.g., c2d
% TODO: relate non-factorable measures to VCSPs, provenance/probabilistic
% databases
% TODO: simplify the introductory explanations (take inspiration from the
% comments of reviewer #5) (this is more about the next two sections)

% Over the last fifteen years, WMC has been extended and generalised in many
% ways, e.g., to handle continuous probability distributions
% \cite{DBLP:conf/ijcai/BellePB15}, first-order probabilistic theories
% \cite{DBLP:conf/ijcai/BroeckTMDR11,DBLP:journals/cacm/GogateD16}, and infinite
% domains \cite{DBLP:conf/aaai/Belle17}.

% , whereas approximate solvers work by sampling
% \cite{DBLP:conf/aaai/ChakrabortyFMSV14} and performing local search
% \cite{DBLP:conf/sat/WeiS05}.

% TODO: rephrase
Weighted model counting (WMC), i.e., an extension of model counting ($\#\SAT{}$)
that assigns a weight to every model \cite{DBLP:conf/aaai/SangBK05}, has emerged
as one of the most dominant and competitive approaches for handling inference
tasks in a wide range of formalisms including Bayesian networks
\cite{DBLP:conf/aaai/SangBK05} and probabilistic programs
\cite{DBLP:journals/tplp/FierensBRSGTJR15}. Furthermore, by generalising the
notion of weights to an arbitrary semiring, a range of other problems are also
captured \cite{DBLP:journals/japll/KimmigBR17}. Exact WMC solvers typically rely
on either knowledge compilation \cite{DBLP:conf/ijcai/LagniezM17} or exhaustive
DPLL search \cite{DBLP:conf/aaai/SangBK05}.

@@

This new measure-theoretic perspective allows us to not only encode any
discrete probability distribution but also improve inference speed. In
\cref{sec:bns}, we demonstrate this by developing a new WMC encoding for
Bayesian networks that uses \emph{conditional weights} on literals (in the
spirit of conditional probabilities) that have literal-based WMC as a special
case. We prove the correctness of the encoding and show how a state-of-the-art
WMC solver ADDMC \cite{DBLP:conf/aaai/DudekPV20} can be adapted to the new
format. ADDMC is a recently-proposed algorithm for WMC based on manipulating
functions on Boolean algebras using an efficient representation for such
functions known as algebraic decision diagrams (ADDs)
\cite{DBLP:journals/fmsd/BaharFGHMPS97}. Finally, in \cref{sec:experiments}, we
show how the new encoding results in faster inference on the vast majority of
benchmark instances, often by one or two orders of magnitude. We explain the
performance benefits by showing how our encoding has asymptotically fewer
variables and ADDs.

%making it applicable to domains such as parsing, dynamic programming, constraint
%programming, databases, and networks.

%conditional weights allow us to represent
%probabilities in a manner that is both efficient and interpretable. Not having
%additional variables that are only needed to manipulate the representation into
%a specific format helps with both efficiency and interpretability. Similarly to
%Bayesian networks, the size of the representation scales with independence.

@@

\paragraph{WMC for Bayesian networks.} Hitherto, four techniques have been
proposed for encoding Bayesian networks into instances of WMC. We will identify
them based on the initials of authors as well as publications years:
\texttt{d02} \cite{DBLP:conf/kr/Darwiche02}, \texttt{sbk05}
\cite{DBLP:conf/aaai/SangBK05}, \texttt{cd05} \cite{DBLP:conf/ijcai/ChaviraD05},
and \texttt{cd06} \cite{DBLP:conf/sat/ChaviraD06}. Below we summarise the
observed performance differences among them. Sang
\etal{}~\shortcite{DBLP:conf/aaai/SangBK05} claim that \texttt{sbk05} is a
smaller encoding than \texttt{d02} with respect to both the number of clauses
and the number of variables but provide no experimental comparison. Chavira and
Darwiche~\shortcite{DBLP:conf/ijcai/ChaviraD05} compare \texttt{cd05} with
\texttt{d02} by measuring the time it takes to compile either encoding into an
arithmetic circuit (but do not measure inference time). The more recent encoding
\texttt{cd05} always compiles faster and results in a smaller arithmetic circuit
(as measured by the number of edges). In their subsequent paper, the same
authors perform two sets of experiments (that are relevant to this summary)
\cite{DBLP:conf/sat/ChaviraD06}. First, they compile \texttt{cd05} and
\texttt{cd06} encodings into d-DNNF (i.e., deterministic decomposable negation
normal form \cite{DBLP:journals/jancl/Darwiche01}), measuring both compilation
time and numbers of edges in the d-DNNF diagram. The results are mostly in
favour of \texttt{cd06}. Second, they compare the inference time of
\texttt{sbk05} run with Cachet \cite{DBLP:conf/sat/SangBBKP04} with the compile
times of \texttt{cd05} and \texttt{cd06}, but only on five (types of) instances.
In these experiments, \texttt{cd06} is always faster than \texttt{cd05}, while
the comparison with \texttt{sbk05} is mixed. Sometimes \texttt{cd06} is orders
of magnitude faster than \texttt{sbk05}, sometimes slightly slower. The
performance difference between \texttt{sbk05} and \texttt{cd05} is even harder
to judge: \texttt{sbk05} is better on three out of five instances and worse on
the remaining two. Based on this description, one would expect \texttt{cd06} to
be faster than both \texttt{cd05} and \texttt{sbk05}, both of which should be
faster than \texttt{d02}. The experiments in \cref{sec:experiments}, however,
strongly disagree with this prediction, showing that the quality of an encoding
depends strongly on the underlying search algorithm or compilation technique.

% TODO: maybe summarise this in one sentence
% \paragraph{ADDs and their use in inference.} ADDs provide an efficient way to
% manipulate functions from a Boolean algebra to any algebraic structure (most
% commonly real numbers) \cite{DBLP:journals/fmsd/BaharFGHMPS97}. They have been
% used to represent value functions in Markov decision processes (MDPs)
% \cite{DBLP:conf/uai/HoeySHB99} and extended to handle first-order MDPs
% \cite{DBLP:journals/ai/SannerB09} and discrete and continuous state MDPs
% \cite{DBLP:conf/uai/SannerDB11}. The latter extension has also been used for
% weighted model integration---an extension of WMC that can handle continuous
% variables \cite{DBLP:conf/ijcai/KolbMSBK18}. For Bayesian networks, ADDS have
% been used to represent each conditional probability table (CPT) as an ADD
% \cite{DBLP:conf/icml/ZhaoMP15} and as a combination of ADDs and arithmetic
% circuits \cite{DBLP:conf/ijcai/ChaviraD07}. ADDs are particularly advantageous
% in this situation because of their ability to fully exploit context-specific
% independence, i.e., the observable structure within a CPT that is not inherited
% from the structure of the Bayesian network \cite{DBLP:conf/uai/BoutilierFGK96}.

\section{Boolean Algebras and Propositional Logic} \label{sec:prelims}

\begin{table*}
  \centering
  \begin{tabular}{lcc}
    \toprule
    Name in logic & Boolean-algebraic notation & Set-theoretic notation \\
    \midrule
    Atoms (elements of $U$) & $a, b$ & $a, b$ \\
    \rowcolor{gray!10} Models (elements of $2^U$) & $\neg a \land \neg b, a \land \neg b, \neg a \land b, a \land b$ & $\emptyset, \{a\}, \{b\}, \{a, b\}$ \\
    & $\top$ & $\{ \emptyset, \{a\}, \{b\}, \{a, b\} \}$ \\
    & $\neg a \lor \neg b, a \to b$ & $\{ \emptyset, \{a\}, \{b\} \}, \{ \emptyset, \{a\}, \{a, b\} \}$ \\
    & $b \to a, a \lor b$ & $\{ \emptyset, \{b\}, \{a, b\} \}, \{ \{a\}, \{b\}, \{a, b\} \}$ \\
    & $\neg b, \neg a, a \leftrightarrow b$ & $\{\emptyset, \{a\}\}, \{\emptyset, \{b\}\}, \{\emptyset, \{a, b\}\}$ \\
    & $(a \land \neg b) \lor (b \land \neg a), a, b$ & $\{\{a\}, \{b\}\}, \{\{a\}, \{a, b\}\}, \{\{b\}, \{a, b\}\}$ \\
    & $\neg a \land \neg b, a \land \neg b, \neg a \land b, a \land b$ & $\{\emptyset\}, \{\{a\}\}, \{\{b\}\}, \{\{a, b\}\}$ \\
    \multirow{-7}{*}{Formulas (elements of $2^{2^U}$)} & $\bot$ & $\emptyset$ \\
    \bottomrule
  \end{tabular}
  \caption{Notation for a logic with two atoms. The elements in both columns are
    listed in the same order.}
  \label{tbl:notation_example}
\end{table*}

@@

% TODO: explain what CUDD is
We also consider the space of all functions from any Boolean algebra to
$\mathbb{R}_{\ge 0}$ together with some operations on those functions. They will
be instrumental in defining WMC as a measure in \cref{sec:wmc_as_measure} and
can be efficiently represented using ADDs. Furthermore, all of the operations
are supported by CUDD package that is used by ADDMC for ADD manipulation
\cite{DBLP:conf/aaai/DudekPV20}. The definitions of multiplication and
projection are as defined by Dudek \etal{}~\shortcite{DBLP:conf/aaai/DudekPV20},
while others are new.

@@

\section{WMC as a Measure on a Boolean Algebra} \label{sec:wmc_as_measure}
% TODO beginning of section 4: a weight function is *any* function; end the
% paragraph with 'for all x \in U'.

% TODO: mention that putting a measure on a BA is a well-known idea
In this section, we introduce an alternative definition of WMC and demonstrate
how it relates to the standard one. Let $U$ be a set. A \emph{measure} is a
function $\mu\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ such that $\mu(\bot) = 0$,
and $\mu(a \lor b) = \mu(a) + \mu(b)$ for all $a, b \in 2^{2^U}$ whenever $a
\land b = \bot$. A \emph{weight function} is a function $\nu\colon 2^U \to
\mathbb{R}_{\ge 0}$. A weight function is \emph{factored} if $\nu = \prod_{x \in
  U} \nu_x$ for some functions $\nu_x\colon 2^{\{x\}} \to \mathbb{R}_{\ge 0}$,
$x \in U$. We say that a weight function $\nu\colon 2^U \to \mathbb{R}_{\ge 0}$
\emph{induces} a measure $\mu_\nu\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ if
$\mu_\nu(x) = \sum_{\{u\} \le x} \nu(u)$.

@@

\paragraph{Relation to the classical (logic-based) view of WMC.} Let
$\mathcal{L}$ be a propositional logic with two atoms $a$ and $b$ as in
\cref{sec:prelims} and $w\colon \{ a, b, \neg a, \neg b \} \to \mathbb{R}_{\ge
  0}$ a weight function defined as $w(a) = 0.3$, $w(\neg a) = 0.7$, $w(b)
= 0.2$, $w(\neg b) = 0.8$. Furthermore, let $\Delta$ be a theory in
$\mathcal{L}$ with a sole axiom $a$. Then $\Delta$ has two models: $\{ a, b \}$
and $\{ a, \neg b \}$ and its WMC \cite{DBLP:journals/ai/ChaviraD08} is
\begin{equation} \label{eq:wmc_example}
  \begin{split}
    \mathrm{WMC}(\Delta) &= \sum_{\omega \models \Delta} \prod_{\omega \models l} w(l) \\
    &= w(a)w(b) + w(a)w(\neg b) = 0.3.
  \end{split}
\end{equation}
Alternatively, we can define $\nu_a\colon 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ as
$\nu_a(\{ a \}) = 0.3$, $\nu_a(\emptyset) = 0.7$ and $\nu_b\colon 2^{\{b\}} \to
\mathbb{R}_{\ge 0}$ as $\nu_b(\{ b \}) = 0.2$, $\nu_b(\emptyset) = 0.8$. Let
$\mu$ be the measure on $2^{2^U}$ induced by $\nu = \nu_a \cdot \nu_b$. Then,
equivalently to \cref{eq:wmc_example}, we can write
\begin{align*}
  \mu(a) &= \nu(\{ a, b \}) + \nu(\{ a \}) \\
         &= \nu_a(\{a\})\nu_b(\{b\}) + \nu_a(\{a\})\nu_b(\emptyset) = 0.3.
\end{align*}
Thus, one can equivalently think of WMC as summing over models of a theory or
over atoms below an element of a Boolean algebra.

@@

\begin{example}
  Let $U = \{a, b\}$ be a set of atoms and $\mu\colon 2^{2^U} \to
  \mathbb{R}_{\ge 0}$ a measure defined as $\mu(a \land b) = 0.72$, $\mu(a \land
  \neg b) = 0.18$, $\mu(\neg a \land b) = 0.07$, $\mu(\neg a \land \neg b) =
  0.03$.\footnote{The value of $\mu$ on any other element of $2^{2^U}$ can be
    deduced from the definition of a measure.} If $\mu$ could be represented
  using literal-weight (factored) WMC, we would have to find two weight
  functions $\nu_a\colon 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ and $\nu_b\colon
  2^{\{b\}} \to \mathbb{R}_{\ge 0}$ such that $\nu = \nu_a \cdot \nu_b$ induces
  $\mu$, i.e., $\nu_a$ and $\nu_b$ would have to satisfy this system of
  equations:
  \begin{align*}
    \nu_a(\{a\}) \cdot \nu_b(\{b\}) &= 0.72 \\
    \nu_a(\{a\}) \cdot \nu_b(\emptyset) &= 0.18 \\
    \nu_a(\emptyset) \cdot \nu_b(\{b\}) &= 0.07 \\
    \nu_a(\emptyset) \cdot \nu_b(\emptyset) &= 0.03,
  \end{align*}
  which has no solution.

@@

Note that in this case, we chose to interpret $\nu_b$ as $\Pr(b \mid a)$
while---with a different definition of $\nu_b$ that represents the joint
probability distribution $\Pr(a, b)$---$\nu_b$ by itself could induce $\mu$. In
general, however, factorising the full weight function into several smaller
functions often results in weight functions with smaller domains which leads to
increased efficiency and decreased memory usage \cite{DBLP:conf/aaai/DudekPV20}.
We can easily generalise this example further.

@@

Since many measures of interest may not be factorable, a well-known way to
encode them into instances of WMC is by adding more literals
\cite{DBLP:journals/ai/ChaviraD08}. We can use the measure-theoretic perspective
on WMC to show that this is always possible, however, as ensuing sections will
demonstrate, it often makes the inference task much harder in
practice.\footnote{The proofs of this and other theoretical results can be found
in the technical appendix.}

@@

\section{Example: Bayesian Networks} \label{sec:bns}
% TODO: make it clear which parts of the encoding are not new
% TODO: rewrite the encoding without using an algorithm (look into the feedback
% of reviewer #5 and someone before him)
% TODO: explain/prove why it's impossible for the all-zero encoding to represent
% the last value (a.k.a. the -implicit flag)

>>>>

\documentclass{uai2021} % for initial submission
% \documentclass[accepted]{uai2021} % after acceptance, for a revised
                                    % version; also before submission to
                                    % see how the non-anonymous paper
                                    % would look like
%% There is a class option to choose the math font
% \documentclass[mathfont=cm]{uai2021} % Computer Modern math instead of
                                       % ptmx, like default for UAI ≤2020
% \documentclass[mathfont=newtx]{uai2021} % newtx fonts (improves upon
                                          % ptmx; less tested, no support)
% NOTE: Only keep *one* line above as appropriate, as it will be replaced
%       automatically for papers to be published. Do not make any other
%       change above this note for an accepted version.

%% Choose your variant of English; be consistent
% \usepackage[american]{babel}
\usepackage[british]{babel}

%% Some suggested packages, as needed:
\usepackage{natbib} % has a nice set of citation styles and commands
    \bibliographystyle{plainnat}
    \renewcommand{\bibsection}{\subsubsection*{References}}
\usepackage{mathtools} % amsmath with fixes and additions
\usepackage[binary-units]{siunitx} % for proper typesetting of numbers and units
\usepackage{booktabs} % commands to create good-looking tables
\usepackage[table]{xcolor}
\usepackage{tikz} % nice language for creating drawings and diagrams

\usepackage{complexity}
\usepackage[capitalise]{cleveref}
\usepackage{multirow}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{thm-restate}
\usepackage{subcaption}
\usepackage[ruled,vlined]{algorithm2e}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

\DeclareMathOperator{\im}{im}

@@

@@

% The standard author block has changed for UAI 2021 to provide
% more space for long author lists and allow for complex affiliations
%
% All author information is authomatically removed by the class for the
% anonymous submission version of your paper, so you can already add your
% information below.
%
% Add authors in order of decreasing contribution
\author[1]{\href{mailto:Harry Q. Bovik <harryq@example.edu>?Subject=Your UAI 2021 paper}{Harry~Q.~Bovik}{}} % Lead author
\author[2]{Coauthor~One}
\author[1,2]{Coauthor~Two}
\author[3]{Further~Coauthor}
\author[1]{Further~Coauthor}
\author[3]{Further~Coauthor}
\author[3,1]{Further~Coauthor}
% Add affiliations after the authors
\affil[1]{%
    Computer Science Dept.\\
    Cranberry University\\
    Pittsburgh, Pennsylvania, USA
}
\affil[2]{%
    Affiliation\\
    Address\\
    …
}
\affil[3]{…}

\begin{document}
\maketitle

@@

\section{Introduction}

Weighted model counting (WMC), i.e., an extension of model counting ($\#\SAT{}$)
that assigns a weight to every model \citep{DBLP:conf/aaai/SangBK05}, has
emerged as one of the most dominant and competitive approaches for handling
inference tasks in a wide range of formalisms including Bayesian networks
\citep{DBLP:conf/aaai/SangBK05,DBLP:books/daglib/0024906}, probabilistic
graphical models more generally \citep{DBLP:conf/ecsqaru/ChoiKD13}, and
probabilistic programs
\citep{DBLP:journals/tplp/FierensBRSGTJR15,DBLP:journals/corr/abs-2005-09089}.
Over the last fifteen years, WMC has been extended and generalised in many ways,
e.g., to handle continuous probability distributions
\citep{DBLP:conf/ijcai/BellePB15}, first-order probabilistic theories
\citep{DBLP:conf/ijcai/BroeckTMDR11,DBLP:journals/cacm/GogateD16}, and infinite
domains \citep{DBLP:conf/aaai/Belle17}. Furthermore, by generalising the notion
of weights to an arbitrary semiring, a range of other problems are also captured
\citep{DBLP:journals/japll/KimmigBR17}. Exact WMC solvers typically rely on
either knowledge compilation
\citep{DBLP:conf/ijcai/OztokD15,DBLP:conf/ijcai/LagniezM17} or exhaustive DPLL
search \citep{DBLP:conf/aaai/SangBK05}, whereas approximate solvers work by
sampling \citep{DBLP:conf/aaai/ChakrabortyFMSV14} and performing local search
\citep{DBLP:conf/sat/WeiS05}.

@@

This new measure-theoretic perspective allows us to not only encode any
discrete probability distribution but also improve inference speed. In
\cref{sec:bns}, we demonstrate this by developing a new WMC encoding for
Bayesian networks that uses \emph{conditional weights} on literals (in the
spirit of conditional probabilities) that have literal-based WMC as a special
case. We prove the correctness of the encoding and show how a state-of-the-art
WMC solver ADDMC \citep{DBLP:conf/aaai/DudekPV20} can be adapted to the new
format. ADDMC is a recently-proposed algorithm for WMC based on manipulating
functions on Boolean algebras using an efficient representation for such
functions known as algebraic decision diagrams (ADDs)
\citep{DBLP:journals/fmsd/BaharFGHMPS97}. Finally, in \cref{sec:experiments}, we
show how the new encoding results in faster inference on the vast majority of
benchmark instances, often by one or two orders of magnitude. We explain the
performance benefits by showing how our encoding has asymptotically fewer
variables and ADDs.

@@

Hitherto, four techniques have been proposed for encoding Bayesian networks into
instances of WMC. We will identify them based on the initials of authors as well
as publications years: \texttt{d02} \citep{DBLP:conf/kr/Darwiche02},
\texttt{sbk05} \citep{DBLP:conf/aaai/SangBK05}, \texttt{cd05}
\citep{DBLP:conf/ijcai/ChaviraD05}, and \texttt{cd06}
\citep{DBLP:conf/sat/ChaviraD06}. Below we summarise the observed performance
differences among them. \citet{DBLP:conf/aaai/SangBK05} claim that
\texttt{sbk05} is a smaller encoding than \texttt{d02} with respect to both the
number of clauses and the number of variables but provide no experimental
comparison. \citet{DBLP:conf/ijcai/ChaviraD05} compare \texttt{cd05} with
\texttt{d02} by measuring the time it takes to compile either encoding into an
arithmetic circuit (but do not measure inference time). The more recent encoding
\texttt{cd05} always compiles faster and results in a smaller arithmetic circuit
(as measured by the number of edges). In their subsequent paper, the same
authors perform two sets of experiments (that are relevant to this summary)
\citep{DBLP:conf/sat/ChaviraD06}. First, they compile \texttt{cd05} and
\texttt{cd06} encodings into d-DNNF (i.e., deterministic decomposable negation
normal form \citep{DBLP:journals/jancl/Darwiche01}), measuring both compilation
time and numbers of edges in the d-DNNF diagram. The results are mostly in
favour of \texttt{cd06}. Second, they compare the inference time of
\texttt{sbk05} run with Cachet \citep{DBLP:conf/sat/SangBBKP04} with the compile
times of \texttt{cd05} and \texttt{cd06}, but only on five (types of) instances.
In these experiments, \texttt{cd06} is always faster than \texttt{cd05}, while
the comparison with \texttt{sbk05} is mixed. Sometimes \texttt{cd06} is orders
of magnitude faster than \texttt{sbk05}, sometimes slightly slower. The
performance difference between \texttt{sbk05} and \texttt{cd05} is even harder
to judge: \texttt{sbk05} is better on three out of five instances and worse on
the remaining two. Based on this description, one would expect \texttt{cd06} to
be faster than both \texttt{cd05} and \texttt{sbk05}, both of which should be
faster than \texttt{d02}. The experiments in \cref{sec:experiments}, however,
strongly disagree with this prediction, showing that the quality of an encoding
depends strongly on the underlying search algorithm or compilation technique.

% \paragraph{ADDs and their use in inference.} ADDs provide an efficient way to
% manipulate functions from a Boolean algebra to any algebraic structure (most
% commonly real numbers) \citep{DBLP:journals/fmsd/BaharFGHMPS97}. They have been
% used to represent value functions in Markov decision processes (MDPs)
% \citep{DBLP:conf/uai/HoeySHB99} and extended to handle first-order MDPs
% \citep{DBLP:journals/ai/SannerB09} and discrete and continuous state MDPs
% \citep{DBLP:conf/uai/SannerDB11}. The latter extension has also been used for
% weighted model integration---an extension of WMC that can handle continuous
% variables \citep{DBLP:conf/ijcai/KolbMSBK18}. For Bayesian networks, ADDS have
% been used to represent each conditional probability table (CPT) as an ADD
% \citep{DBLP:conf/icml/ZhaoMP15} and as a combination of ADDs and arithmetic
% circuits \citep{DBLP:conf/ijcai/ChaviraD07}. ADDs are particularly advantageous
% in this situation because of their ability to fully exploit context-specific
% independence, i.e., the observable structure within a CPT that is not inherited
% from the structure of the Bayesian network \citep{DBLP:conf/uai/BoutilierFGK96}.

\section{Boolean Algebras, Power Sets, and Propositional
  Logic} \label{sec:prelims}

\begin{table*}
  \caption{Notation for a logic with two atoms. The elements in both columns are
    listed in the same order.}
  \label{tbl:notation_example}
  \centering
  \begin{tabular}{lcc}
    \toprule
    Name in logic & Boolean-algebraic notation & Set-theoretic notation \\
    \midrule
    Atoms (elements of $U$) & $a, b$ & $a, b$ \\
    \rowcolor{gray!10} Models (elements of $2^U$) & $\neg a \land \neg b, a \land \neg b, \neg a \land b, a \land b$ & $\emptyset, \{a\}, \{b\}, \{a, b\}$ \\
    & $\top$ & $\{ \emptyset, \{a\}, \{b\}, \{a, b\} \}$ \\
    & $\neg a \lor \neg b, a \to b$ & $\{ \emptyset, \{a\}, \{b\} \}, \{ \emptyset, \{a\}, \{a, b\} \}$ \\
    & $b \to a, a \lor b$ & $\{ \emptyset, \{b\}, \{a, b\} \}, \{ \{a\}, \{b\}, \{a, b\} \}$ \\
    & $\neg b, \neg a, a \leftrightarrow b$ & $\{\emptyset, \{a\}\}, \{\emptyset, \{b\}\}, \{\emptyset, \{a, b\}\}$ \\
    & $(a \land \neg b) \lor (b \land \neg a), a, b$ & $\{\{a\}, \{b\}\}, \{\{a\}, \{a, b\}\}, \{\{b\}, \{a, b\}\}$ \\
    & $\neg a \land \neg b, a \land \neg b, \neg a \land b, a \land b$ & $\{\emptyset\}, \{\{a\}\}, \{\{b\}\}, \{\{a, b\}\}$ \\
    \multirow{-7}{*}{Formulas (elements of $2^{2^U}$)} & $\bot$ & $\emptyset$ \\
    \bottomrule
  \end{tabular}
\end{table*}

@@

We also consider the space of all functions from any Boolean algebra to
$\mathbb{R}_{\ge 0}$ together with some operations on those functions. They will
be instrumental in defining WMC as a measure in \cref{sec:wmc_as_measure} and
can be efficiently represented using ADDs. Furthermore, all of the operations
are supported by the CUDD \citep{somenzi1998cudd} package that is used by ADDMC
for ADD manipulation \citep{DBLP:conf/aaai/DudekPV20}. The definitions of
multiplication and projection are as defined by
\citet{DBLP:conf/aaai/DudekPV20}, while others are new.

@@

\section{WMC as a Measure on a Boolean Algebra} \label{sec:wmc_as_measure}

In this section, we introduce an alternative definition of WMC and demonstrate
how it relates to the standard one. Let $U$ be a set. A \emph{measure} is a
function $\mu\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ such that $\mu(\bot) = 0$,
and $\mu(a \lor b) = \mu(a) + \mu(b)$ for all $a, b \in 2^{2^U}$ whenever $a
\land b = \bot$ \citep{gaifman1964concerning,DBLP:books/daglib/0090259}. A
\emph{weight function} is a function $\nu\colon 2^U \to \mathbb{R}_{\ge 0}$. A
weight function is \emph{factored} if $\nu = \prod_{x \in U} \nu_x$ for some
functions $\nu_x\colon 2^{\{x\}} \to \mathbb{R}_{\ge 0}$, $x \in U$. We say that
a weight function $\nu\colon 2^U \to \mathbb{R}_{\ge 0}$ \emph{induces} a
measure $\mu_\nu\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ if $\mu_\nu(x) =
\sum_{\{u\} \le x} \nu(u)$.

@@

\paragraph{Relation to the classical (logic-based) view of WMC.} Let
$\mathcal{L}$ be a propositional logic with two atoms $a$ and $b$ as in
\cref{sec:prelims} and $w\colon \{ a, b, \neg a, \neg b \} \to \mathbb{R}_{\ge
  0}$ a weight function defined as $w(a) = 0.3$, $w(\neg a) = 0.7$, $w(b)
= 0.2$, $w(\neg b) = 0.8$. Furthermore, let $\Delta$ be a theory in
$\mathcal{L}$ with a sole axiom $a$. Then $\Delta$ has two models: $\{ a, b \}$
and $\{ a, \neg b \}$ and its WMC \citep{DBLP:journals/ai/ChaviraD08} is
\begin{equation} \label{eq:wmc_example}
  \begin{split}
    \mathrm{WMC}(\Delta) &= \sum_{\omega \models \Delta} \prod_{\omega \models l} w(l) \\
    &= w(a)w(b) + w(a)w(\neg b) = 0.3.
  \end{split}
\end{equation}
Alternatively, we can define $\nu_a\colon 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ as
$\nu_a(\{ a \}) = 0.3$, $\nu_a(\emptyset) = 0.7$ and $\nu_b\colon 2^{\{b\}} \to
\mathbb{R}_{\ge 0}$ as $\nu_b(\{ b \}) = 0.2$, $\nu_b(\emptyset) = 0.8$. Let
$\mu$ be the measure on $2^{2^U}$ induced by $\nu = \nu_a \cdot \nu_b$. Then,
equivalently to \cref{eq:wmc_example}, we can write
\begin{align*}
  \mu(a) &= \nu(\{ a, b \}) + \nu(\{ a \}) \\
         &= \nu_a(\{a\})\nu_b(\{b\}) + \nu_a(\{a\})\nu_b(\emptyset) = 0.3.
\end{align*}
Thus, one can equivalently think of WMC as summing over models of a theory or
over atoms below an element of a Boolean algebra.

@@

\begin{example}
  Let $U = \{a, b\}$ be a set of atoms and $\mu\colon 2^{2^U} \to
  \mathbb{R}_{\ge 0}$ a measure defined as $\mu(a \land b) = 0.72$, $\mu(a \land
  \neg b) = 0.18$, $\mu(\neg a \land b) = 0.07$, $\mu(\neg a \land \neg b) =
  0.03$.\footnote{The value of $\mu$ on any other element of $2^{2^U}$ can be
    deduced from the definition of a measure.} If $\mu$ could be represented
  using literal-weight (factored) WMC, we would have to find two weight
  functions $\nu_a\colon 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ and $\nu_b\colon
  2^{\{b\}} \to \mathbb{R}_{\ge 0}$ such that $\nu = \nu_a \cdot \nu_b$ induces
  $\mu$, i.e., $\nu_a$ and $\nu_b$ would have to satisfy this system of
  equations:
  \begin{align*}
    \nu_a(\{a\}) \cdot \nu_b(\{b\}) &= 0.72 \\
    \nu_a(\{a\}) \cdot \nu_b(\emptyset) &= 0.18 \\
    \nu_a(\emptyset) \cdot \nu_b(\{b\}) &= 0.07 \\
    \nu_a(\emptyset) \cdot \nu_b(\emptyset) &= 0.03,
  \end{align*}
  which has no solutions.

@@

Note that in this case, we chose to interpret $\nu_b$ as $\Pr(b \mid a)$
while---with a different definition of $\nu_b$ that represents the joint
probability distribution $\Pr(a, b)$---$\nu_b$ by itself could induce $\mu$. In
general, however, factorising the full weight function into several smaller
functions often results in weight functions with smaller domains which leads to
increased efficiency and decreased memory usage
\citep{DBLP:conf/aaai/DudekPV20}. We can easily generalise this example further.

@@

Since many measures of interest may not be factorable, a well-known way to
encode them into instances of WMC is by adding more literals
\citep{DBLP:journals/ai/ChaviraD08}. We can use the measure-theoretic
perspective on WMC to show that this is always possible, however, as ensuing
sections will demonstrate, it often makes the inference task much harder in
practice.\footnote{The proofs of this and other theoretical results can be found
  in the technical appendix.}

@@

\section{Encoding Bayesian Networks Using Conditional Weights} \label{sec:bns}
